Constraining the High Density Nuclear Symmetry Energy with Pions

### 2020-07-14 chapters/ch_data_analysis_2.tex
Bill's edits and mine to results

\chapter{Data Analysis II}

\section{Radio Isotope Beam Factory (RIBF) Facility }
\label{sec:beam}
%Cyclotron facility overview.
%Samurai line overview.
%Beam line element overview.
%Big rips beam PID. reference 
The primary and secondary beams were produced at the Radioactive Isotope Beam Factory (RIFB) facility at RIKEN, in Wako-shi, Japan. The RIBF facility starts with two primary beam types, ${}^{132}$Xe and ${}^{238}$U, which are produced by an ion-source and accelerated to progressively higher kinetic energies by a linear accelerator (RILAC), and 4 different cyclotrons (RRC, fRC, IRC, and SRC), until they reach a beam energy of \SI{345}{\MeVA}. Figure~\ref{fig:samuraiBeamLine} shows the later stages of the cyclotrons and the following beam lines they feed into.

\begin{figure}[!htb]
\includegraphics[width=\linewidth]{SAMURAI-beamline.png}
\caption{Overview of the RIBF, BigRIPS, and SAMURAI beamline.}
\label{fig:samuraiBeamLine}
\end{figure}

After the SCR, the primary beams impinge on a rotating \SI{3}{\milli\metre} Be target which produces many different species by fragmentation. These fragments are then separated by the BigRIPS spectrometer which is tuned to the particular secondary fragment of interest. This is accomplished through several dipole magnets, slits, and wedge degraders. The resulting secondary beam is not pure and the purity depends on the capability of BigRIPS to deliver the secondary beam of choice and the primary beam used.

In these set of experiments several beams were produced with varying intensities and purities. Table~\ref{tb:beams} summarizes the average qualities of the 4 secondary beams produced in the two experimental campaigns where most beams were delivered with an intensity of \SI{10}{\kilo\hertz}.

 \begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}ccccc@{}}\toprule 
 Primary Beam & Secondary Beam & Energy at mid target \si{\MeVA} & Intensity \si{\kilo\hertz} & Purity (\%) \\ [0.5ex] 
 \midrule
 ${}^{238}$U   & ${}^{132}$Sn   &  269.2  &  9.5  &  54   \\
 ${}^{238}$U   & ${}^{124}$Sn   &  270.3  &  9.1  &  10  \\
 ${}^{124}$Xe  & ${}^{112}$Sn   &  270.4  &  7.6  &  48  \\
 ${}^{124}$Xe  & ${}^{108}$Sn   &  269.3  &  7.5  &  52   \\
 \bottomrule
\end{tabular}
\caption{Primary and secondary beam properties produced in the \spirit TPC experimental campaigns. }
\label{tb:beams}
\end{table*}

@@

>>>>

\chapter{Data Analysis II}
As Chapter~\ref{chap:data1} covered the corrections and various calibrations to the data, this Chapter will cover the various cuts taken to improve the signal in the experimental data. Some cuts which identify the incoming beam and the event on target are the minimum cut in order to perform an analysis. The cuts pertaining to the track reconstruction quality are also required to suppress as much of the background noise as possible, most of which comes from reconstruction inefficiencies which the cuts presented here address.

@@

\subsection{Correcting Pion Spectra (Efficiency and Acceptance)}

To correct for the efficiency, we apply a track-by-track correction, where the efficiency of the i-th track $\epsilon_i$ is retrieved from the efficiency database, from the parameters discussed in Section~\ref{sec:efficiency}. The correction factor $C_i$ is defined as,

\begin{equation}
C_i = \epsilon_i^{-1}.
\end{equation}

Each track that is identified as a pion as described in Section~\ref{sec:pid}, is then weighted by the correction factor $C_i$, when filling the histogram of any observable. In this way, we can correct for the efficiency on a track-by-track basis, allowing for any transformation of the track of interest into any observable, notably transforming from the lab frame into the center-of-mass (CM) frame. Before performing the transformation to the CM system, recall there is a small, but non-negligible, beam angle in the Lab frame discussed in Section~\ref{sec:beamangle}. First we rotate each event so that the beam will align with the z-axis. Doing so makes the transformation into the CM system much simpler to describe. If the beam direction is defined by a unit vector $\hat{b}$, we can define the rotation that rotates the beam into the z-axis as a rotation about an arbitrary vector $\hat{v} = \hat{b}\times\hat{z}$ where the angle between the two is given by $\cos \theta = \hat{b}\cdot\hat{z}$. A rotation with the angle $\theta$ is then applied around the vector $\hat{v}$.

Once all the events have been rotated to align with the z-axis, transforming from the Lab to the CM frame is done by a Lorentz transformation. The 4-momentum vector in the lab frame is defined as $\textbf{P} = (E/c,p_x,p_y,p_z)$. Where the corresponding Lorentz transform into the CM frame along the beam (z-axis) is defined as the Lorentz transformation matrix $A$,

\begin{equation}
A = \begin{pmatrix}
1 & 0 & 0 & 0\\
0 & 1 & 0 & 0\\
0 & 0 & \gamma & -\beta \gamma\\
0 & 0 & -\beta \gamma & \gamma
\end{pmatrix},
\end{equation}

where $\beta$, describes the velocity of the CM system, and $\gamma=\sqrt{1-\beta^2}^{-1}$. The parameter $\beta$ can be determined from the total momentum of the system in the Laboratory frame $P = \sqrt{ T_{P}^2 + 2M_{T}T_{T}}$ and the total energy of the system $E = T_{P} + M_{P} + M_{T}$ -- where $T_{P}$ is the projectile kinetic energy in the Lab frame, $M_{T}$, and $M_{P}$ are the total mass of the target and projectile, where $\beta = -P/E$; the (-) sign denotes the correct direction for transforming from the Lab to the CM frame. The transformation of each  track is defined as $p^{CM} = \textbf{A}p^{Lab}$.

### 2020-07-14 chapters/ch_experiment.tex
Bill's edits and mine to results

%explain in breif the basics of a time projection chamber 
\begin{figure}[!htb]
\centering
\includegraphics[scale=.5]{tpcPrinciple.pdf}
\caption{Operation principle of the TPC}
\label{fig:tpcPrinciple}
\end{figure}

@@

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{exploded.png}
\caption{Overview of the \spirit TPC}
\label{fig:tpcExplode}
\end{figure}

@@

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}rr@{}}\toprule 
\multicolumn{2}{c}{\spirit TPC Overview} \\
 \midrule
Pad plane area & 1.3 m x .9 m\\
Pad size       & 1.2 cm x .8 cm \\
Number of pads & 12096 (112 x 108) \\
Gas composition& 90\% Ar + 10\% CH${}_4$ (1 atm)  \\
Multiplicity limit & 200  \\
dE/dx range        & Z=1-3, $\pi$, p, d, t, He, Li \\
Drift length       & 50 cm \\
\bottomrule
\end{tabular}
\caption{An overview of the properties of the \spirit TPC}
\label{tb:spiritoverview}
\end{table*}

@@

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{fc_overview.png}
\label{fig:fc_overview}
\caption{FC overview}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{fc_overview2.png}
\label{fig:fc_overview2}
\caption{FC overview 2}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{TPC_schematic.pdf}
\caption{Schematic of the TPC system}
\label{fig:TPC_schematic}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[scale=.6]{FC_schematic.pdf}
\caption{Schematic of the electric connections relevant to the Field Cage system. The strip thickness is exaggerated in the figure to show the detail}
\label{fig:FC_schematic}
\end{figure}

@@

 \begin{table*}[!htb]
 \centering
\ra{1.3}
\begin{tabular}{@{}rrrrrrr@{}}\toprule 
\multicolumn{2}{c}{GET electronics settings}\\
\midrule
 Plane & Material & Diameter \si{\micro\metre} & Pitch \si{\milli\metre} & Distance to pad-plane & Tension \si{\newton} & Voltage \si{\volt}\\ [0.5ex] 
 Anode  & Au-plated W   &  20  &  4  &  4   &  0.5  &  1460  \\
 Ground & BeCu          &  75  &  1  &  8   &  1.2  &  0     \\ 
 Gating & BeCu          &  75  &  1  &  14   &  1.2 &  -110$\pm$ 70\\ 
 \bottomrule
\end{tabular}
\caption{Wire plane properties}
\label{tb:wireplane}
\end{table*}

@@

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{wires_open.pdf} 
        \caption{Wires open} \label{fig:wires_open}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.42\textwidth}
        \centering
        \includegraphics[width=\linewidth]{wires_closed.pdf} 
        \caption{Wires closed.} \label{fig:wires_closed}
    \end{subfigure}
\label{fig:wires}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{gg_onoff.pdf}
\caption{On and off configurations of the gating grid.}
\label{fig:gg_onoff}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{averageTransparency.png} 
        \caption{Electron transparency for the conditions of all wires are the same voltage.} \label{fig:ggAvgTrans}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{transparencyDeltaV.png} 
        \caption{Electrons transparency for the mode where adjacent wires have a voltage difference of $2 \Delta V$.} \label{fig:ggDeltaVTrans}
    \end{subfigure}
\label{fig:ggTrans}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=.5\textwidth]{GET_and_ZAP.pdf}
\caption{•}
\label{fig:getzap}
\end{figure}

@@

The average energy loss in a material can be described by the Bethe-Bloch equation,
\begin{equation}\label{eq:bb}
\frac{dE}{dx} = \frac{4\pi NZ^2e^4}{mc^2\beta^2} (ln \frac{2mc^2\beta^2\gamma^2}{I} - \beta^2),
\end{equation}
where $N$ is the number density of electrons in the medium, $e$ the elementary charge, $mc^2$ is the rest mass of the electron, $Z$ is the charge of the traversing particle, $I$ is the mean excitation energy of the medium, and $\beta$ is the velocity of the particle \cite{blumrol}. Yet there is a large variation in energy loss around this mean value. The statistical variation of energy loss in a material was described by Landau \cite{landau} and later better described by Shulek \cite{shulek} and Bichsel \cite{bichsel1}. In both approximations it is described by a most probable energy loss value, with a long, high-energy loss tail. The solid curve in Fig.~\ref{fig:straggling} shows the energy loss distribution in Ar gas for a proton with momentum \SI{3.4}{\giga\eVperc} \cite{bichsel}. The dashed line is the distribution under the Landau assumptions. The mean energy loss $\langle\Delta\rangle$ is significantly shifted from the most probable value $\Delta_p$, due to the long high energy tail.  Because of this long tail, for a finite set of energy loss measurements along a given track, the mean value is a very unreliable. The most probable energy loss is the better observable which can be obtained either through fitting of the observed distribution or through the truncated mean method.

\begin{figure}
\includegraphics[width=\textwidth]{bichsel.png}
\caption{Energy loss of a $\beta\gamma$ = 3.6 particle in Ar gas taken from \cite{bichsel}}
\label{fig:straggling}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=\linewidth]{driftvel.png}
\caption{Drift velocity of electrons in P10 gas.}
\label{fig:driftvel}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{DataTrigger1.png} 
        \caption{${}^{124}$Xe primary beam trigger.} \label{fig:dataTrigger1}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{DataTrigger2.png} 
        \caption{${}^{238}$U primary beam trigger.} \label{fig:dataTrigger2}
    \end{subfigure}
\label{fig:datatrigger}
\end{figure}

>>>>

%explain in breif the basics of a time projection chamber 
\begin{figure}[!htb]
\centering
\includegraphics[scale=.5]{tpcPrinciple.pdf}
\caption{Cartoon depicting the operating principles of the TPC is.}
\label{fig:tpcPrinciple}
\end{figure}

@@

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{exploded.png}
\caption{Exploded overview of the of the \spirit TPC}
\label{fig:tpcExplode}
\end{figure}

@@

\begin{table*}\centering
\ra{1.3}
\begin{tabular}{@{}rr@{}}\toprule 
\multicolumn{2}{c}{\spirit TPC Overview} \\
 \midrule
Pad plane area & 1.3 m x .9 m\\
Pad size       & 1.2 cm x .8 cm \\
Number of pads & 12096 (112 x 108) \\
Gas composition& 90\% Ar + 10\% CH${}_4$ (1 atm)  \\
Multiplicity limit & 200  \\
dE/dx range        & Z=1-3, $\pi$, p, d, t, He, Li \\
Drift length       & 50 cm \\
\bottomrule
\end{tabular}
\caption{Summary of general properties of the \spirit TPC.}
\label{tb:spiritoverview}
\end{table*}

@@

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{fc_overview.png}
\label{fig:fc_overview}
\caption{Overview of the field cage mechanical components.}
\end{figure}

\begin{figure}[!htb]
\includegraphics[width=\textwidth]{fc_overview2.png}
\label{fig:fc_overview2}
\caption{Schematic of the field cage.}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{TPC_schematic.pdf}
\caption{Schematic of the resistive elements in the TPC system.}
\label{fig:TPC_schematic}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[scale=.6]{FC_schematic.pdf}
\caption{Schematic of the electric connections relevant to the field cage system. The strip thickness is exaggerated in the figure to show the detail.}
\label{fig:FC_schematic}
\end{figure}

@@

 \begin{table*}[!htb]
 \centering
\ra{1.3}
\begin{tabular}{@{}rrrrrrr@{}}\toprule 
\multicolumn{2}{c}{GET electronics settings}\\
\midrule
 Plane & Material & Diameter \si{\micro\metre} & Pitch \si{\milli\metre} & Distance to pad-plane & Tension \si{\newton} & Voltage \si{\volt}\\ [0.5ex] 
 Anode  & Au-plated W   &  20  &  4  &  4   &  0.5  &  1460  \\
 Ground & BeCu          &  75  &  1  &  8   &  1.2  &  0     \\ 
 Gating & BeCu          &  75  &  1  &  14   &  1.2 &  -110$\pm$ 70\\ 
 \bottomrule
\end{tabular}
\caption{Wire plane properties.}
\label{tb:wireplane}
\end{table*}

@@

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{wires_open.pdf} 
        \caption{Wires open} \label{fig:wires_open}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.42\textwidth}
        \centering
        \includegraphics[width=\linewidth]{wires_closed.pdf} 
        \caption{Wires closed.} \label{fig:wires_closed}
    \end{subfigure}
\caption{Cartoon depiction of electron drift lines going through the various wire planes.}
\label{fig:wires}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{gg_onoff.pdf}
\caption{Garfield simulations of the on and off configuration of the gating grid.}
\label{fig:gg_onoff}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{averageTransparency.png} 
        \caption{On configurations where all wires are the same voltage.} \label{fig:ggAvgTrans}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.49\textwidth}
        \centering
        \includegraphics[width=\linewidth]{transparencyDeltaV.png} 
        \caption{Off configuration where adjacent wires have a voltage difference of $2 \Delta V$.} \label{fig:ggDeltaVTrans}
    \end{subfigure}

\caption{Electron transparency as measured and simulated in the on and off configurations of the gating grid.}    
\label{fig:ggTrans}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=.5\textwidth]{GET_and_ZAP.pdf}
\caption{Drawing of the ZAP and AsAd board which is connected to the surface pads on the backside of the pad-plane.}
\label{fig:getzap}
\end{figure}

@@

The average energy loss in a material can be described by the Bethe-Bloch equation,
\begin{equation}\label{eq:bb}
\frac{dE}{dx} = \frac{4\pi NZ^2e^4}{mc^2\beta^2} (\ln \frac{2mc^2\beta^2\gamma^2}{I} - \beta^2),
\end{equation}
where $N$ is the number density of electrons in the medium, $e$ the elementary charge, $mc^2$ is the rest mass of the electron, $Z$ is the charge of the traversing particle, $I$ is the mean excitation energy of the medium, and $\beta$ is the velocity of the particle \cite{blumrol}. Yet there is a large variation in energy loss around this mean value. The statistical variation of energy loss in a material was described by Landau \cite{landau} and later better described by Shulek \cite{shulek} and Bichsel \cite{bichsel1}. In both approximations it is described by a most probable energy loss value, with a long, high-energy loss tail. The solid curve in Fig.~\ref{fig:straggling} shows the energy loss distribution in Ar gas for a proton with momentum \SI{3.4}{\giga\eVperc} \cite{bichsel}. The dashed line is the distribution under the Landau assumptions. The mean energy loss $\langle\Delta\rangle$ is significantly shifted from the most probable value $\Delta_p$, due to the long high energy tail.  Because of this long tail, for a finite set of energy loss measurements along a given track, the mean value is a very unreliable. The most probable energy loss is the better observable which can be obtained either through fitting of the observed distribution or through the truncated mean method.

\begin{figure}
\includegraphics[width=\textwidth]{bichsel.png}
\caption{Energy loss distribution of a $\beta\gamma$ = 3.6 particle in Ar gas taken from \cite{bichsel}}
\label{fig:straggling}
\end{figure}

@@

\begin{figure}[H]
\includegraphics[width=\linewidth]{driftvel.png}
\caption{Drift velocity of electrons in P10 gas as a function of the reduced electric field.}
\label{fig:driftvel}
\end{figure}

@@

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{DataTrigger1.png} 
        \caption{${}^{124}$Xe primary beam trigger.} \label{fig:dataTrigger1}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=\linewidth]{DataTrigger2.png} 
        \caption{${}^{238}$U primary beam trigger.} \label{fig:dataTrigger2}
    \end{subfigure}
\caption{Master triggers in each primary beam setting.}    
\label{fig:datatrigger}
\end{figure}

### 2020-07-14 chapters/ch_intro.tex
Bill's edits and mine to results

\chapter{Introduction}
Mankind has been interested in what is the nature of the matter which composes the visible universe, and what are the constituents that make it up. The most significant discovery on the study of fundamental sub-atomic particles was Rutherford's discovery in 1911 in which he concluded that atomic nuclei are composed of a dense nucleus surrounded by an electron cloud. Later it was discovered that the nucleus is composed of nucleons, a positive charge (protons) and a neutral charge (neutrons). The nucleus itself accounts for 99.9\% of the mass of the atom and only \num{e-12} of the total volume, making it incredibly dense.

@@

Large macroscopic objects such as neutron stars are composed of mostly pure neutron matter CITE HERE with a typical size of \SI{11}{\kilo\metre} in radius. The di-neutron system is typically not stable in such an extreme system except for the the large gravitational force, which balances the strong forces. In this extreme example of dense nuclear matter, the pressures generates in the neutron star range from low densities near the crust to the dense interior which can reach up to 9$\rho_o$ CITE HERE. To understand these exotic forms of nuclear matter, the energy density of the system must be described in a more general way than Eq.~\ref{eq:semiEmp}, where we must describe nuclear matter over a wide range of densities.

@@

\section{Pion Observable}
It is preferable to find an observable that is easier to measure experimentally than neutrons, and is more sensitive to the just the high density region. Pions are produced though an intermediate step via $ NN \leftrightarrow \Delta$, where nucleon-nucleon collisions form an excited $\Delta$(1232) baryon resonance from one of the nucleons, which then decay decay shortly after into a pion $\Delta \leftrightarrow \pi N$. The threshold for $\Delta$ resonance production, with a mass of \SI{1232}{\mega\electronvolt\per\clight\squared}, corresponds to a laboratory beam of \SI{290}{\MeVA} in a fixed target experiment. In large nuclei, the internal motion of the nucleons is substantial and non-negligible. In the Fermi gas model, nucleons are arranged filling up higher energy levels up until they reach the Fermi energy. This extra energy allows for $\Delta$ production even at sub-threshold beam energies \cite{fermiEnergy}.

>>>>

\chapter{Introduction}
On August 17, 2017, a collision of two presumptive neutron stars produced gravitational waves which were measured by the LIGO and Advanced Virgo detectors, in combination with several other Astronomical telescopes, marking a new era of advanced multi-messenger observations, reinvigorating the question about the nature of dense matter in the universe.

Mankind has been interested in what is the nature of the matter which composes the visible universe, and what are the constituents that make it up. The most significant discovery on the study of fundamental sub-atomic particles was Rutherford's discovery in 1911 in which he concluded that atomic nuclei are composed of a dense nucleus surrounded by an electron cloud. Later it was discovered that the nucleus is composed of nucleons, a positive charge (protons) and a neutral charge (neutrons). The nucleus itself accounts for 99.9\% of the mass of the atom and only \num{e-12} of the total volume, making it incredibly dense.

@@

Large macroscopic objects such as neutron stars are composed of mostly pure neutron matter \cite{neutronstar} with a typical size of \SI{11}{\kilo\metre} in radius. The di-neutron system is typically not stable in such an extreme system except for the the large gravitational force, which balances the strong forces. In this extreme example of dense nuclear matter, the pressures generates in the neutron star range from low densities near the crust to the dense interior which can reach up to 9$\rho_o$ \cite{neutronstar}. To understand these exotic forms of nuclear matter, the energy density of the system must be described in a more general way than Eq.~\ref{eq:semiEmp}, where we must describe nuclear matter over a wide range of densities.

@@

\section{Pion Observable}
\label{sec:pionObs}
It is preferable to find an observable that is easier to measure experimentally than neutrons, and is more sensitive to the just the high density region. Pions are produced though an intermediate step via $ NN \leftrightarrow \Delta$, where nucleon-nucleon collisions form an excited $\Delta$(1232) baryon resonance from one of the nucleons, which then decay decay shortly after into a pion $\Delta \leftrightarrow \pi N$. The threshold for $\Delta$ resonance production, with a mass of \SI{1232}{\mega\electronvolt\per\clight\squared}, corresponds to a laboratory beam of \SI{290}{\MeVA} in a fixed target experiment. In large nuclei, the internal motion of the nucleons is substantial and non-negligible. In the Fermi gas model, nucleons are arranged filling up higher energy levels up until they reach the Fermi energy. This extra energy allows for $\Delta$ production even at sub-threshold beam energies \cite{fermiEnergy}.
